Rice yield improvements by plant breeding are offset by decreases over time

### 2017-05-10 output/breed_paper.Rnw
Updates paper to use RES data back to 1984

@@

<<echo = FALSE>>=

library(rstan)
library(rstanarm)
library(lme4)

source("../src/prep_model_data.R")
load("h1_fit.Rda")

@@

California (CA) rice production represents one such rice production
system. Rice yields in CA during the period from 1920 -- 1990
increased rapidly, but since 1990 have increased more slowly (Fig. 1
\ref{fig:state_yield}). However, continued improvement of grain yield
as the result of plant breeding has been demonstrated since 1990
\citep{rrb}, with each new cultivar having a yield advantage over the
previous cultivars. Yet, these same reports show stagnant yields; for
example, at the year of release (1985), cultivar 'M-202' averaged
11,060 kg ha\textsuperscript{-1} \citep[14\% MC;][]{m202}. In 2015,
the average yield of the newest released cultivar 'M-209' was 10,841
kg ha\textsuperscript{-1} \citep{m209}. This suggests little to no
progress via breeding, however 'M-209' demonstrated an overall 10\%
yield advantage over 'M-202' in side-by-side comparisons \citep{m209}. We sought to
better understand how grain yields have increased in the breeders'
fields without a corresponding increase in yield in the farmers'
fields. Two competing hypotheses were tested: (1) Grain yields have
not improved since 1990, or (2) grain yield improvements have been
offset by an associated decrease in yield once a cultivar was
selected.

\begin{figure}[ht]
<<fig=TRUE, echo=FALSE>>=
CA_data = read.csv("../data/CA_rice_yield.csv", stringsAsFactors = FALSE)
CA_data$Value = as.numeric(gsub(",", "", CA_data$Value))
CA_data = CA_data[order(CA_data$Year),]
plot(Value * 1.12 ~ Year, CA_data,
     # type = "l",
     pch = 16,
     xlab = "Year",
     ylab = expression("Average grain yield ( kg ha"^-1~")"))
mv_avg = filter(c(rep(mean(CA_data$Value[1:10] * 1.12), 5),
                  CA_data$Value * 1.12,
                  rep(mean(CA_data$Value[(nrow(CA_data) - 10):nrow(CA_data)] * 1.12), 5)),
                rep(1/11,11))
lines(x = c(1907:1911, CA_data$Year, 2017:2021), y = mv_avg)

@@

In the course of investigating the drivers of yield in this system, we
were compelled to explore how yields could increase with each release
of improved cultivars, yet average yield for the region remain
flat. Specifically, we offered two competing hypotheses to explain
this phenomenon: (1) the yield advantage offered by newly developed
cultivars is not greater than previous cultivars, or (2) yield
increases gained through plant breeding are being offset by decreases
in yield of released cultivars.

Yet, annual reports from the CA Cooperative Rice Research Foundation
show that each released cultivar has a yield advantage over the
previous cultivars, typically on the order of 4 -- 6\%. This raises the
question of how to reconcile these facts. Namely, there are two
possibilities:

@@

The Rice Experiment Station (RES) of the California Cooperative Rice
Research Foundation (CCRRF) is located northwest of Biggs, CA (39.4648,
-121.7342), and has been the central location for the development of
improved rice varieties suited for CA since 1969. The climate at the
RES is Mediterranian, characterized by mild winters during which most
of the annual precipitation occurs (\Sexpr{18*24.5} -- \Sexpr{25*24.5}
mm annually) and warm summers largely free of precipitation
events. Soils at the RES are classified as Esquon-Neerdobe clays with
roughly 2\% SOM \citep{ssurgo} in the top 15cm. As the central
location for the breeding program, the RES is researcher-managed (as
opposed to farmer-managed trials) and conducts the bulk of the trials
for the Statewide Variety Trials (approximately 30\% of the total
plots are planted at RES). For these reasons, we focused our
investigation on the RES to avoid complications due to site, climate, and
management differences at the other 6 -- 8 farmer-managed trials.

As part of the annual Statewide Variety Trials, newly developed
entries are tested against named and released cultivars (checks) at
the RES in multiple trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as medium grain cultivars constitute roughly
90\% of the planted area in CA. This lead to a final data set of
\Sexpr{format(nrow(mod_data), big.mark = ',')} observations with
\Sexpr{length(unique(mod_data[,'id']))} cultivars represented,
constituting every publicly released medium grain variety developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of release (typically 3 -- 5 years after introduction to the
trials but prior to widespread commercial seed availability) and the
time elapse (in years) since release was calculated.

@@

\begin{equation}
 yield_i = \alpha + \beta_{years\ since\ release\ i}* x_{years\ since\
   release\ i} + \beta_{year\ i} + \beta_{cultivar\ i}
\end{equation}

Where $yield_i$ is the grain yield for plot $i$, $\alpha$ is the
average yield for the RES across varieties and years (intercept),
$\beta_{years\ since\ release\ i}$ is effect of $x_{years\ since\
release\ i}$, the number of years $i$'s cultivar has been in the
trials (negative for years prior to release, zero at the year of
release), $\beta_{year}$ is the random effect of year, and
$\beta_{cultivar}$ is the random effect of cultivar. A similar but
separate model was used to estimate the yield advantage of newer
releases over older ones:

\begin{equation}
yield_i = \alpha + \beta_{release\ year\ i}*x_{release\ year\ i} +
\beta_{year\ i} + \beta_{cultivar\ i}
\end{equation}

Where $\beta_{release\ year\ i}$ is effect of $x_{release\ year\ i}$,
the year of release for $i$'s cultivar. All other parameters are
similar to Eq. 1.

@@

<<echo=FALSE>>=
rel_CI = round(posterior_interval(RES_rel_fit, pars = "release_yr_c") * scl, 1)

posterior_interval(RES_rel_fit, pars = "release_yr_c") * scl
m202_v_206 = ((posterior_predict(RES_rel_fit, newdata = data.frame(release_yr_c = c(-15, 4)),
                                 re.form = NA) * scl) + center)
m202_v_206 = m202_v_206[,2] - m202_v_206[,1]
m202_v_206 = round(quantile(m202_v_206, p = c(0.05, 0.5, 0.95)), 1)
@

There was a positive relationship between the year of release and
yield, with an estimated \Sexpr{round(fixef(RES_rel_fit)[2] * scl, 1)}
kg ha\textsuperscript{-1} year\textsuperscript{-1} increase (90\% CI:
\Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]})
(Fig. \ref{fig:release}). Comparing the two dominate cultivars M-202,
which was widely planted from the late 1980s until the early 2000s,
and M-206, currently the most widely planted CRRFS cultivar
(approximately 50\% of planted area), there is an estimated
\Sexpr{m202_v_206[2]} kg ha\textsuperscript{-1} yield advantage of M-206
over M-202 (Figs \ref{fig:release} \& \ref{fig:compare_yrs}).

@@

<<fig=TRUE, echo=FALSE>>=
yrs <- 1980:2015
pp <- posterior_predict(RES_rel_fit, re.form = NA, newdata = data.frame(release_yr_c = yrs - 2000))
pp <-  (pp * scl + center)

ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))

res = mod_data$site == "RES"
plot_data = prep_plot_data(mod_data$yield_kg[res] ,
                           mod_data$id[res])

@@

plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(8000, 12500),
     xlim = c(1980,2019))
abline(a = (((fixef(RES_rel_fit)[1] * scl) + center) ) -
           (2000 * fixef(RES_rel_fit)[2] * scl),
       b = fixef(RES_rel_fit)[2] * scl, col = "grey50",
       lty = 2)
with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
text(x = plot_data$release_yr, y = plot_data$yield_kg,
     labels = gsub("M", "M-", row.names(plot_data)),
     pos = 4, cex = 0.8)

text(1990, 8000, bquote(.(round(fixef(RES_rel_fit)[2] * scl , 1))~kg~ha^-1~year^-1))

@@

\caption{The yield increase over time of medium grain rice varieties
  in California, USA. There is an estimated
  increase in yield over time, but it is smaller than a previously
  estimate of 54 lbs year\textsuperscript{-1}. This estimate is
  potentially misleading, because it only
  considers the average yield of each variety during the period 1995
  -- 2015, and the older varieties have experiencing greater losses
  since being released.}
\label{fig:release}
\end{figure}

@@

plot_data = lapply(tars, function(x) {
    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
    tmp = prep_plot_data(dd$yield_kg , dd$year)
    tmp$year = as.numeric(row.names(tmp))
    tmp
    })

invisible(sapply(1:2, function(i){
    plot(mean ~ year,
         data = plot_data[[i]],
         xlim = c(1995, 2016), ylim = c(7500,13500),
         pch = 16, main = gsub("M","M-", tars[i]),
         ylab = "",
         yaxt = "n",
         xlab = "Years after release")
    segments(x0 = plot_data[[i]]$year,
             y0 = plot_data[[i]]$mean + plot_data[[i]]$se,
             y1 = plot_data[[i]]$mean - plot_data[[i]]$se)
    if(i == 1){
        axis(2)
        mtext(expression("Yield ( kg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
    }

    abline(h = mean(mod_data$yield_kg[(mod_data$id == tars[i] &
                                       mod_data$site == "RES")] ),
           lty = 1, col = "grey50")

}))

@
\caption{The performance of two major cultivars at the CA Rice
  Experiment Station (RES) over the period 1995 -- 2015. The line is the average
  yield for each cultivar over the period which it was planted at the
  RES. The newer cultivar, M-206, appears to have a slight yield
  advantage (an estimated \Sexpr{m202_v_206[2]} kg ha\textsuperscript{-1})over the older cultivar, M-202.}
\label{fig:compare_yrs}
\end{figure}

<<echo=FALSE>>=
all_CI <- round(posterior_interval(RES_yrs_fit, pars = "yrs_in_trial") * scl, 1)
RES_CI <- round(posterior_interval(RES_yrs_fit, pars = "yrs_in_trial") * scl, 1)
@

When yields are arranged according to the years since release, there
is a small decrease in yield over time once a cultivar is released,
though the estimated effect is small compared to the variability in
the data set. This decrease is estimated to be
\Sexpr{round(fixef(RES_yrs_fit)[2] * scl, 1)} kg
ha\textsuperscript{-1} year\textsuperscript{-1} (90\% CI: \Sexpr{RES_CI[2]} --
\Sexpr{RES_CI[1]}). The oldest cultivar
in the trials is estimated to have lost
\Sexpr{round(fixef(RES_yrs_fit)[2] * scl * 36 , 1)} kg
ha\textsuperscript{-1} since release.

Again comparing the yield of two prominant cultivars, M-202 and M-206,
arranged not by the year but rather the years since release, we find
that the two cultivars are compariable and that the estimated yield
advantage (\Sexpr{m202_v_206[2]} kg ha\textsuperscript{-1}) can be
explained by the differences in the amount of time that has passed
since each cultivar was released
(Fig. \ref{fig:compare_since_release})

<<fig=TRUE, echo=FALSE>>=
yrs_in <- -7:35
pp <- posterior_predict(RES_yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
pp <-  (pp * scl + center)

RES = res = mod_data$site == "RES"
plot_data = prep_plot_data(mod_data$yield_kg[res] ,
                           mod_data$yrs_in_trial[res])

plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Year since release",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(8000, 12500))
with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
abline(a = ((fixef(RES_yrs_fit)[1] * scl) + center) ,
       b = fixef(RES_yrs_fit)[2] * scl, col = "grey50",
       lty = 2)
ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
text(5, 8000, bquote(.(round(fixef(RES_yrs_fit)[2] * scl , 1))~kg~ha^-1~year^-1))

@
\caption{The relationship between yield and the years since release for
  medium grain rice varieties planted at the Rice Experiment Station
  (RES), California, USA. RES is considered
  the best managed site, and has the most plots per year of any of the
  sites. The estimated decrease over time is estimated to be stronger
  using only these data (\Sexpr{round(fixef(RES_yrs_fit)[2] * scl, 1)}
  lbs year\textsuperscript{-1}), primarly due to lower variability.}
\label{fig:RES_year}
\end{figure}

\begin{figure}

plot_data = lapply(tars, function(x) {
    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
    tmp = prep_plot_data(dd$yield_kg , dd$yrs_in_trial)
    tmp$yrs_in_trial = as.numeric(row.names(tmp))
    tmp
    })

plot(mean ~ yrs_in_trial,
     data = plot_data[[1]],
     xlim = c(-6, 32), ylim = c(6500,14500),
     pch = 16, main = gsub("M","M-", tars[i]),
     ylab = expression("Yield ( kg ha"^-1~")"),
     xlab = "Years after release")
segments(x0 = plot_data[[1]]$yrs_in_trial,
         y0 = plot_data[[1]]$mean + plot_data[[1]]$se,
         y1 = plot_data[[1]]$mean - plot_data[[1]]$se)
points(mean ~ yrs_in_trial,
       data = plot_data[[2]],
       pch = 17, col = "grey40")
segments(x0 = plot_data[[2]]$yrs_in_trial,
         y0 = plot_data[[2]]$mean + plot_data[[2]]$se,
         y1 = plot_data[[2]]$mean - plot_data[[2]]$se,
         col = "grey40")
abline(lm(yield_kg ~ yrs_in_trial,
          data = mod_data[(mod_data$id %in% tars & mod_data$site == "RES"),]),
       lty = 1, col = "grey50")
legend("topright", legend = gsub("M","M-", tars),
       pch = c(16, 17), col = c("black","grey40"))

Contrary to the estimates for decreases once a cultivar is released, there was a
small increase over time (Fig. \ref{fig:release}), though it was small
(\Sexpr{round(fixef(RES_rel_fit)[2] * scl, 1)} lbs
year\textsuperscript{-1}). This suggests that the overall progress on
yield over the period from 1980 -- 2016 is approximately
\Sexpr{round(fixef(RES_rel_fit)[2] * scl * 36, 1)} lbs acre\textsuperscript{-1}.

@@

There is evidence of a small decrease in yield over time once a cultivar is
released. This complicates comparisons between new entries and checks,
since the check will decrease over time. This decrease will manifest
as an apparent larger yield advantage of the new entry over the check.
There is evidence of a small increase in yield with each new released
cultivar. Therefore, it appears that successive cultivar releases are
not merely recovering to the old benchmark but are surpassing
it. However, progress is slower than previously estimated
(\Sexpr{round(fixef(RES_rel_fit)[2] * scl, 1)} lbs
acre\textsuperscript{-1} year\textsuperscript{-1} increase verses a
previous estimate of 54 lbs acre\textsuperscript{-1}
year\textsuperscript{-1}).

@@

\newpage
\section{References}\label{references}

@@

\begin{figure}[h]

<<label=plot1, fig=TRUE, echo=FALSE>>=
boxplot(yield_kg ~ year, data = mod_data,
        xlab = "Year", ylab = expression("Yield ( lb ac"^-1~")"))
@
  \caption{Distribution of yield over the period 1995 -- 2015.} \label{fig:plot}
\end{figure}

\begin{figure}[h]
<<label=plot2, fig=TRUE, echo=FALSE>>=

  boxplot(yield_kg ~ id, data = mod_data,
          xlab = "Cultivar",
          ylab = "Yield (lbs/ac)")
@
\caption{Boxplot of yields for released medium grain cultivars for
  the period 1995--2015.}
\label{fig:raw_yield}
\end{figure}

>>>>

RES: Rice Experiment Station

CCRRF: California Cooperative Rice Research Foundation

@@

<<echo=FALSE>>=

library(rstan)
library(rstanarm)
library(lme4)
library(xtable)

load("../data/model_data.rda")
load("h1_fit.Rda")
load("../data/all_vt_weather.Rda")

@@

California (CA) rice production represents one such rice production
system. Rice yields in CA during the period from 1920 -- 1990
increased rapidly, in part due to improvements in rice genetics (i.e.,
the Green Revolution) through the efforts of the California
Cooperative Rice Research Foundation (CCRRF; a collaboration between
the University of California, the USDA-Agricultural Research Service,
and farmer-funded research). However, since 1990 CA state average
yields have increased more slowly (Fig. \ref{fig:state_yield}) despite
demonstrated improvements of grain yield as the result of plant
breeding \citep{ccrrf}, with each new cultivar having a yield
advantage over the previous cultivars in side-by-side comparisons in
the annual Statewide Variety Trials. Yet, taken across years, these
same reports show stagnant yields; for example, at the year of release
(1985), cultivar M-202 averaged 11 Mg ha\textsuperscript{-1}
\citep[14\% MC;][]{m202}. In 2015, the average yield of the newest
released cultivar M-209 was 10.8 Mg ha\textsuperscript{-1}
\citep{m209}. In other systems, it has been shown that rice cultivars
experience a lack of adaptation to changing biotic and abiotic
conditions, resulting in a decrease of a cultivar's performance over
time \citep{Peng2010, Peng2000, Datta1995}. We sought to explore this
phenomonon and better understand how grain yields have increased in
the breeders' fields without a corresponding increase in yield in the
farmers' fields. Two competing hypotheses were tested: (1) the yield
advantage offered by newly developed cultivars is not greater than
previous cultivars, or (2) yield increases gained through plant
breeding are being offset by decreases in the yield of released
cultivars over time.

\begin{figure}[ht]
<<fig=TRUE, echo=FALSE>>=
CA_data = read.csv("../data/CA_rice_yield.csv", stringsAsFactors = FALSE)
CA_data$Value = as.numeric(gsub(",", "", CA_data$Value))/1000
CA_data = CA_data[order(CA_data$Year),]
plot(Value * 1.12 ~ Year, CA_data,
     # type = "l",
     pch = 16,
     xlab = "Year",
     ylab = expression("Average grain yield ( Mg ha"^-1~")"))
mv_avg = filter(c(rep(mean(CA_data$Value[1:10] * 1.12), 5),
                  CA_data$Value * 1.12,
                  rep(mean(CA_data$Value[(nrow(CA_data) - 10):nrow(CA_data)] * 1.12), 5)),
                rep(1/11,11))
lines(x = c(1907:1911, CA_data$Year, 2017:2021), y = mv_avg)

@@

@@

The Rice Experiment Station (RES) of the CCRRF is located northwest of
Biggs, CA (39.4648, -121.7342), and has been the central location for
the development of improved rice varieties adapted for CA since
1969. The climate at the RES is Mediterranian, characterized by mild
winters during which most of the annual precipitation occurs
(\Sexpr{18*24.5} -- \Sexpr{25*24.5} mm annually) and warm summers
largely free of precipitation events. Soils at the RES are classified
as Esquon-Neerdobe clays with roughly 2\% SOM \citep{ssurgo} in the
top 15cm. As the central location for the breeding program, the RES is
researcher-managed (as opposed to farmer-managed trials) and conducts
the bulk of the trials for the Statewide Variety Trials (approximately
3 -- 4 times the number of plots compared to any other location). For
these reasons, we focused our investigation on the RES to avoid
potential complications due to site, climate, and management
differences at the other 6 -- 8 farmer-managed trials per year.

As part of the annual Statewide Variety Trials, newly developed
entries are tested against named and released cultivars (checks) at
the RES in multiple trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as medium grain cultivars constitute roughly
90\% of the planted area in CA. For these cultivars planted in
breeding trials at the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations were
collected. This lead to a final data set of
\Sexpr{format(nrow(mod_data), big.mark = ',')} observations with
\Sexpr{length(unique(mod_data[,'id']))} cultivars represented,
including every publicly released medium grain variety developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of release (typically 3 -- 5 years after introduction to the
trials but prior to widespread commercial seed availability) and the
time elapse (in years) since release was calculated.

Plots at the RES are managed similar to perscribed best management
practice, including land preparation, fertility management, and pest
and disease control. For planting, rice seed is pre-germinated prior
to direct-seeding into pre-flooded fields. Plots were either 14.0 or
18.5 m\textsuperscript{2}, and were harvested after physiological
maturity using a small-plot combine. All yields were converted to 14\%
moisture content prior to reporting.

@@

\begin{equation}
 yield_i = \alpha + \beta_{years\ since\ release\ i}* x_{years\ since\
   release\ i} + \beta_{release\ year\ i}*x_{release\ year\ i} + \beta_{year\ i} + \beta_{cultivar\ i}
\end{equation}

Where $yield_i$ is the grain yield for plot $i$, $\alpha$ is the
average yield for the RES across varieties and years (intercept),
$\beta_{years\ since\ release\ i}$ is effect of $x_{years\ since\
release\ i}$, the number of years $i$'s cultivar has been in the
trials (negative for years prior to release, zero at the year of
release), $\beta_{release\ year\ i}$ is the effect of $x_{release\
year\ i}$, the year of release for $i$'s cultivar, $\beta_{year}$ is
the random effect of year, and $\beta_{cultivar}$ is the random effect
of cultivar.

@@

<<echo=FALSE>>=
rel_CI = round(posterior_interval(RES_full, pars = "release_yr_c") * scl, 1)

m202_v_206 = round(fixef(RES_full)[2] * scl * 19, 0)
# m202_v_206 = round((posterior_interval(RES_full, pars = "release_yr_c") * scl) * 19, 0)

# m202_v_206 = ((posterior_predict(RES_full, newdata = data.frame(release_yr_c = c(-15, 4)),
#                                  re.form = NA) * scl) + center)
# m202_v_206 = m202_v_206[,2] - m202_v_206[,1]
# m202_v_206 = round(quantile(m202_v_206, p = c(0.05, 0.5, 0.95)), 1)

There was a positive relationship between the year of release and
yield, with an estimated \Sexpr{round(fixef(RES_full)[2] * scl, 1)}
kg ha\textsuperscript{-1} year\textsuperscript{-1} increase (90\% CI:
\Sexpr{rel_CI[1]} -- \Sexpr{rel_CI[2]})
(Fig. \ref{fig:release}). Comparing the two dominate cultivars M-202,
which was widely planted from the late 1980s until the early 2000s,
and M-206, currently the most widely planted CRRFS cultivar
(approximately 50\% of planted area), there is an estimated
\Sexpr{m202_v_206} kg ha\textsuperscript{-1} yield advantage of
M-206 over M-202 (Figs \ref{fig:release} \& \ref{fig:compare_yrs}).

@@

<<fig=TRUE, echo=FALSE>>=
yrs_in <- -10:40
pp <- posterior_linpred(RES_full, re.form = NA,
                        newdata = data.frame(yrs_in_trial = yrs_in,
                                             release_yr_c = 0))
pp <-  (pp * scl + center) / 1000 
ints = apply(pp, 2, quantile, p = c(0.05, 0.95))

RES = res = mod_data$site == "RES"
plot_data = prep_plot_data(mod_data$yield_kg[res] / 1000,
                           mod_data$yrs_in_trial[res])

plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Year since release",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(8.0, 12.5))
with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 ,
       b = fixef(RES_full)[3] * scl/1000)
polygon(x = c(yrs_in, rev(yrs_in)),
        y = c(ints[1,],rev(ints[2,])),
        border = NA,
        col = rgb(0,0,0,0.1))
text(5, 8.0, bquote(.(round(fixef(RES_full)[3] * scl , 1))~kg~ha^-1~year^-1))

@
\caption{The relationship between yield and the years since release
  for medium grain rice varieties planted at the Rice Experiment Station
  (RES), California, USA. The estimated decrease over time once a
  cultivar is release is estimated to be
  \Sexpr{round(fixef(RES_full)[3] * scl, 1)} kg ha\textsuperscript{-1}
  year\textsuperscript{-1}. The shaded area is the 90\% credible
  interval.}
\label{fig:RES_year}
\end{figure}

\begin{figure}[ht]
<<fig=TRUE, echo=FALSE>>=
yrs <- 1975:2025
pp <- posterior_linpred(RES_full, re.form = NA,
                        newdata = data.frame(release_yr_c = yrs - 2000,
                                             yrs_in_trial = 0),
                        prob = 0.5)
pp <-  (pp * scl + center) / 1000
ints = apply(pp, 2, quantile, p = c(0.05, 0.95))

res = mod_data$site == "RES"
plot_data = prep_plot_data(mod_data$yield_kg[res]/1000 ,
                           mod_data$id[res])

@@

plot(yield_kg ~ release_yr, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = expression("Grain yield ( Mg ha"^-1~")"),
     ylim = c(8, 12.5),
     xlim = c(1980,2019))
abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
           (2000 * fixef(RES_full)[2] * scl)) / 1000,
       b = fixef(RES_full)[2] * scl/1000)
abline(lm(yield_kg/1000 ~ release_yr, data = mod_data[res,]), lty = 2)
with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
text(x = plot_data$release_yr, y = plot_data$yield_kg + 0.15,
     labels = gsub("M", "M-", row.names(plot_data)),
     pos = 4, cex = 0.8)

legend("bottomright", legend = c("w/ years since release",
                                 "w/o years since release",
                                 "90% Credible interval"),
       lty = c(1,2, NA),
       bty = "n",
       fill = c(NA,NA, "grey"),
       border = NA,
       merge = TRUE)
polygon(x = c(yrs, rev(yrs)),
        y = c(ints[1,],rev(ints[2,])),
        border = NA,
        col = rgb(0,0,0,0.1))

# text(1990, 8.0, bquote(.(round(fixef(RES_full)[2] * scl , 1))~kg~ha^-1~year^-1))

@@

\caption{The yield increase over time of medium grain rice varieties
  in California, USA. Ignoring the years since release (dashed line)
  suggests an increase in yield potential over time. However, correcting
  the comparison to account for differences in the length of time since
  each cultivar was released (solid line) shows no increase in yield over time.}
\label{fig:release}
\end{figure}

@@

plot_data = lapply(tars, function(x) {
    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
    tmp = prep_plot_data(dd$yield_kg/1000 , dd$year)
    tmp$year = as.numeric(row.names(tmp))
    tmp
    })

invisible(sapply(1:2, function(i){
    plot(mean ~ year,
         data = plot_data[[i]],
         xlim = c(1985, 2016), ylim = c(7.5,13.5),
         pch = 16, main = gsub("M","M-", tars[i]),
         ylab = "",
         yaxt = "n",
         xlab = "Year",
         cex.axis = 0.8)
    segments(x0 = plot_data[[i]]$year,
             y0 = plot_data[[i]]$mean + plot_data[[i]]$se,
             y1 = plot_data[[i]]$mean - plot_data[[i]]$se)
    if(i == 1){
        axis(2, cex.axis = 0.8,)
        mtext(expression("Yield ( Mg ha"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
    }

    # abline(h = mean(mod_data$yield_kg[(mod_data$id == tars[i] &
    #                                    mod_data$site == "RES")] ),
    #        lty = 1, col = "grey50")
    mod = lm(mean ~ year, data = plot_data[[i]])
    tmp = predict(mod)
    if(i == 1){
        abline(mod)
    }else{
        lines(as.numeric(names(tmp)), tmp)
    }
    abline(h = 11.2, lty = 2)
    text(2000, 7.5, bquote(.(round(coef(mod)[2] * 1000, 1))~kg~ha^-1~year^-1))
}))

@
\caption{The performance of two major cultivars at the CA Rice
  Experiment Station (RES) over the period 1995 -- 2015. The solid line
  represents a simple linear regression of yield for each cultivar over
  the period which it was planted at the RES (including only that
  cultivar's data), while the dashed line is the average yield of M-202
  at its year of release (1985). Both cultivars have a similar yield at
  their respective year of release, and similar decreases over
  time. The slope for the individual regressions are comparible to the
slope estimated from all medium grain cultivars
(\Sexpr{round(fixef(RES_full)[3] * scl, 1)} kg ha\textsuperscript{-1} year\textsuperscript{-1}).}
\label{fig:compare_yrs}
\end{figure}

<<echo=FALSE>>=
all_CI <- round(posterior_interval(RES_full, pars = "yrs_in_trial") * scl, 1)
RES_CI <- round(posterior_interval(RES_full, pars = "yrs_in_trial") * scl, 1)
@

When yields are arranged according to the years since release, there
is a small decrease in yield over time once a cultivar is released,
though the estimated effect is small compared to the variability in
the data set. This decrease is estimated to be
\Sexpr{round(fixef(RES_full)[3] * scl, 1)} kg
ha\textsuperscript{-1} year\textsuperscript{-1} (90\% CI: \Sexpr{RES_CI[2]} --
\Sexpr{RES_CI[1]}). The oldest cultivar
in the trials is estimated to have lost
\Sexpr{round(fixef(RES_full)[3] * scl * 36 , 1)} kg
ha\textsuperscript{-1} since release.

Again comparing the yield of two prominant cultivars, M-202 and M-206,
arranged not by the year but rather the years since release, we find
that the two cultivars are compariable and that the estimated yield
advantage (\Sexpr{m202_v_206} kg ha\textsuperscript{-1}) can be
explained by the differences in the amount of time that has passed
since each cultivar was released
(Fig. \ref{fig:compare_yrs} \& \ref{fig:compare_since_release})

Contrary to the estimates for decreases once a cultivar is released,
there was no increase in yield potential over time once the
differences between the years since release were accounted for (Fig. \ref{fig:release})
(\Sexpr{round(fixef(RES_full)[2] * scl, 1)} kg ha\textsuperscript{-1}
year\textsuperscript{-1}). This suggests that the overall progress on
yield over the period from 1980 -- 2016 is approximately
\Sexpr{round(fixef(RES_full)[3] * scl * 36, 1)} lbs acre\textsuperscript{-1}.

@@

 We postulate that true
yield increases are only possible if the improvement in yield of each
new cultivar exceeds former's benchmark yield at the time of release,
and therefore side-by-side comparisons are potentially misleading.

There is evidence of a small decrease in yield over time once a cultivar is
released. This complicates comparisons between new entries and checks,
since the check will decrease over time. This decrease will manifest
as an apparent larger yield advantage of the new entry over the check.
There is evidence of a small increase in yield with each new released
cultivar. Therefore, it appears that successive cultivar releases are
not merely recovering to the old benchmark but are surpassing
it. However, progress is slower than previously estimated
(\Sexpr{round(fixef(RES_full)[2] * scl, 1)} lbs
acre\textsuperscript{-1} year\textsuperscript{-1} increase verses a
previous estimate of 54 lbs acre\textsuperscript{-1}
year\textsuperscript{-1}).

@@

\section{Tables}

<<echo=FALSE, results=tex>>=
CIs = posterior_interval(RES_full)[1:3,] * scl
CIs[1,] = round(CIs[1,] + center, -2)
CIs = round(CIs, 1)
med_eff = fixef(RES_full) * scl
med_eff[1] = round(med_eff[1] + center, -2)
med_eff = round(med_eff, 1)

RES_coef = cbind(med_eff, paste(CIs[,1],CIs[,2], sep = " -- "))
rownames(RES_coef) = c("Intercept (kg ha\\textsuperscript{-1})",
                       "Release year (kg ha\\textsuperscript{-1} year\\textsuperscript{-1})",
                       "Years since release (kg ha\\textsuperscript{-1} year\\textsuperscript{-1})")
colnames(RES_coef) = c("Median", "90\\% Credible interval")

print(xtable(RES_coef,
             caption = "Estimated median effects and 90\\% credible intervals.",
             label = "tbl:coef",
             align = "lcc"),
      caption.placement="top",
      booktabs=TRUE,
      sanitize.colnames.function = function(x){x},
      sanitize.rownames.function = function(x){x})
@
\newpage
\section{References}\label{references}

@@

plot_data = lapply(tars, function(x) {
    dd = mod_data[mod_data$id == x & mod_data$site == "RES",]
    tmp = prep_plot_data(dd$yield_kg / 1000, dd$yrs_in_trial)
    tmp$yrs_in_trial = as.numeric(row.names(tmp))
    tmp
    })

plot(mean ~ yrs_in_trial,
     data = plot_data[[1]],
     xlim = c(-6, 32), ylim = c(6.5,14.5),
     pch = 16, #main = gsub("M","M-", tars[i]),
     ylab = expression("Yield ( kg ha"^-1~")"),
     xlab = "Years after release")
segments(x0 = plot_data[[1]]$yrs_in_trial,
         y0 = plot_data[[1]]$mean + plot_data[[1]]$se,
         y1 = plot_data[[1]]$mean - plot_data[[1]]$se)
points(mean ~ yrs_in_trial,
       data = plot_data[[2]],
       pch = 17, col = "grey40")
segments(x0 = plot_data[[2]]$yrs_in_trial,
         y0 = plot_data[[2]]$mean + plot_data[[2]]$se,
         y1 = plot_data[[2]]$mean - plot_data[[2]]$se,
         col = "grey40")
abline(lm(yield_kg/1000 ~ yrs_in_trial,
          data = mod_data[(mod_data$id %in% tars & mod_data$site == "RES"),]),
       lty = 1, col = "grey50")
legend("topright", legend = gsub("M","M-", tars),
       pch = c(16, 17), col = c("black","grey40"))

\begin{figure}[ht]
<<fig=TRUE, echo=FALSE>>=
tars <- c("M202","M206")

boxplot(yield_kg/1000 ~ id, data = mod_data[mod_data$id %in% tars,],
        ylab = expression("Grain yield ( Mg ha"^-1~")"))

@
\caption{The performance of two major cultivars at the CA Rice
Experiment Station (RES) over the period 1995 -- 2015. The newer
cultivar, M-206, appears to have a slight yield advantage (an
estimated \Sexpr{m202_v_206} kg ha\textsuperscript{-1}) over the older
cultivar, M-202, though this comparison does not account for the 10
year difference in years since release between the cultivars.}
\label{fig:compare_vars}
\end{figure}

% \begin{figure}[h]

% <<label=plot1, fig=TRUE, echo=FALSE>>=
% boxplot(yield_kg ~ year, data = mod_data,
%         xlab = "Year", ylab = expression("Yield ( kg ha"^-1~")"))
% @
%   \caption{Distribution of yield over the period 1995 -- 2015.} \label{fig:plot}
% \end{figure}

% \begin{figure}[h]
% <<label=plot2, fig=TRUE, echo=FALSE>>=
